Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-07-05 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Ch 2
Addressed CF comments Ch 2 including revised figs

Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs). These spread sheets incorporate references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study because an insufficient amount of reliable data is available to conduct validation.  The correlations for flame height and time to ignition were not included in the validation study.

@@

The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA:921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a ``guide'' as opposed to a ``standard,'' it is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

@@

Plots of the centerline temperatures averaged over the replicate experiments for each of the fuels are presented in Fig.~\ref{Temp}. The natural gas temperatures were averaged over a ``steady-state'' burning period.  The temperature differences between the vertical thermocouple locations, for locations greater than 20~cm above the fire location, begin to decrease as the distances from the top of the burner increase.  This trend is similar for both the gasoline fires and the polyurethane foam fires as well.  The most notable difference between the natural gas and the other fuels is the growth and decay of the fire during the test period.

@@

Plots of the total heat flux for each of the fuels are presented in Fig.~\ref{Flux}. The repeatability of the total heat flux values varied in a manner similar to the heat release rates, in that the natural gas fires provided a steady state value while the gasoline and the polyurethane foam fires generated a fire curve that has periods of growth and decay.   The sensing surface of the heat flux gauges were located 0.95~m horizontally from the center of the burner/fuel and 0.50~m above the top of the burner/pan lip/fuel surface.  The average peak total heat flux for the natural gas fires were approximately 2.0~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 3.5~kW/m$^2$ and 2.5~kW/m$^2$ respectively.  The values are only presented for the experiments which had two heat flux gauges installed as described in the instrumentation section above.

@@

\begin{figure}
  \includegraphics[width=2.7in]{../Figures/Fig14a}
  \includegraphics[width=2.7in]{../Figures/Fig14b} \\
  \caption[Photographs of the natural gas and polyurethane foam fires]{Photographs showing a natural gas fire (left) and a polyurethane foam fire (right) under the oxygen consumption calorimeter.}
  \label{Gas_Foam_Photos}
\end{figure}

\begin{figure}
  \includegraphics[width=2.7in]{../Figures/Fig15a}
  \includegraphics[width=2.7in]{../Figures/Fig15b} \\
  \caption[Photographs of the gasoline fire]{Photograph showing a gasoline fueled experiment with a flame height of 0.65~m (left) and a photograph taken approximately 0.2~s later with a flame height of 1.0~m (right).}
  \label{Gasoline_Photos}
\end{figure}

@@

Figure~\ref{Foam_Contours} shows two different contours for the polyurethane foam fueled experiment.  The results from polyurethane fueled fire characterization experiments have an increased amount of variability due to the nature of the ignition source and the fuel itself.  The polyurethane foam was readily ignited with a small pilot flame similar to that used to ignite the natural gas and the gasoline.  By comparison, a small flame ignited the gases flowing out the top of the natural gas burner and the flames from the natural gas cover entire burner surface within approximately one second.  A similar ignition sequence occurs with the gasoline pan fires. The polyurethane foam has a more complicated flame spread sequence that resulted in significant changes to the peak heat release rate and the time to peak heat release.  Once the polyurethane foam was ignited, it took approximately 60~s for the flame to spread across the top surface of the fuel.  As the flames spread horizontally across the fuel surface, the fire also burns down into the fuel.  In some cases the fire burned completely through the fuel in the center of the foam sample, resulting in a hollow flame at the time of the peak heat release rate.  An example of a contour plot, exhibiting the hollow or bifurcated flames is shown at right in Fig.~\ref{Foam_Contours}.  Other methods of ignition for the polyurethane foam were considered in order to have a more uniform ignition of the top surface of the foam.  Each method would have resulted in a larger release of energy from the ignition source or a starter fuel.   Based on a few scoping experiments these methods did not improve the repeatability.

@@

Once the characteristics of the fires had been measured, the next step was to measure the thermal impact that the fires had on a vertical target surface.  In these experiments the target is a non-combustible wall. The non-combustible material was chosen to minimize any contribution from the wall to the fire.  The fires were placed at predetermined, fixed positions away from the wall.  In the remainder of the report the term burner will be considered synonymous with the pan for the liquid fuel or the block of solid fuel. Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and then touching the wall (0.0~m).

@@

A free standing steel frame wall with a 12.7~mm thick, 2.44~m high by 1.22~m wide, calcium silicate board face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux. Marinite Type I is the commercial name of the non-combustible board used in these experiments.  It is normally used as a fire-resistant thermal insulation that also has structural strength. 90~\% of the board is composed of calcium silicate and calcium meta-silicate.  The remaining 10~\% of the board is composed of a combination of organic fibers, fiber glass filament and crystalline silica (quartz)~\cite{BNZ:MSDS}.  Material properties that impact heat transfer were selected from the manufacturer's literature and are presented in the table below~\cite{Marinite:1997}.

>>>>

\subsection{The Use of Fire Models in Fire Investigation}
Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs). These spread sheets incorporate references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study because an insufficient amount of reliable data is available to conduct validation.  The correlations for flame height and time to ignition were not included in the validation study.

@@

\subsection{Fire Investigation Under Review}
The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA:921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a ``guide'' as opposed to a ``standard,'' it is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

@@

Plots of the centerline temperatures averaged over the full set of replicate experiments for each of the fuels are presented in Fig.~\ref{Temp}. The natural gas temperatures are averaged over a ``steady-state'' burning period.  The temperature differences between the vertical thermocouple locations, for locations greater than 20~cm above the fire location, begin to decrease as the distances from the top of the burner increase.  This trend is similar for both the gasoline fires and the polyurethane foam fires as well.  The most notable difference between the natural gas and the other fuels is the growth and decay of the fire during the test period.

@@

Plots of the total heat flux for each of the fuels are presented in Fig.~\ref{Flux}. The repeatability of the total heat flux values varied in a manner similar to the heat release rates, in that the natural gas fires provided a steady state value while the gasoline and the polyurethane foam fires generated a fire curve that has periods of growth and decay.  The average peak total heat flux for the natural gas fires were approximately 2.0~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 3.5~kW/m$^2$ and 2.5~kW/m$^2$ respectively.  The values are only presented for the experiments which had two heat flux gauges installed as described in the instrumentation section above.

@@

\begin{figure}
  \includegraphics[height=3.5in, width=1.7in]{../Figures/NG_STFire}
  \includegraphics[height=3.5in, width=1.7in]{../Figures/Gaso_STFire} 
  \includegraphics[height=3.5in, width=1.7in]{../Figures/PUF_STFire} \\
  \caption[Photographs of the natural gas, gasoline, and polyurethane foam fires]{Photographs showing a natural gas fire (left), gasoline fire (center), and a polyurethane foam fire (right) under the oxygen consumption calorimeter.}
  \label{Gas_Foam_Photos}
\end{figure}

\begin{figure}
  \includegraphics[width=2.6in]{../Figures/Gaso_0_65m}
  \includegraphics[width=2.6in]{../Figures/Gaso_1_0m} \\
  \caption[Photographs of the gasoline fire]{Photograph showing a gasoline fueled experiment with a flame height of 0.65~m (left) and a photograph taken approximately 0.2~s later with a flame height of 1.0~m (right).}
  \label{Gasoline_Photos}
\end{figure}

@@

Figure~\ref{Foam_Contours} shows two different contours for the polyurethane foam fueled experiment.  The results from polyurethane fueled fire characterization experiments have an increased amount of variability due to the nature of the ignition source and the fuel itself.  The polyurethane foam was readily ignited with a small pilot flame similar to that used to ignite the natural gas and the gasoline.  By comparison, a small flame ignited the gases flowing out the top of the natural gas burner and the flames from the natural gas cover entire burner surface within approximately one second.  A similar ignition sequence occurs with the gasoline pan fires. The polyurethane foam has a more complicated flame spread sequence that resulted in significant changes to the peak heat release rate and the time to peak heat release.  Once the polyurethane foam was ignited, it took approximately 60~s for the flame to spread across the top surface of the fuel.  As the flames spread horizontally across the fuel surface, the fire also burns down into the fuel.  In some cases the fire burned completely through the fuel in the center of the foam sample, resulting in a hollow flame at the time of the peak heat release rate.  An example of a contour plot, exhibiting the hollow or bifurcated flames is shown at right in Figure~\ref{Foam_Contours}. Refer back to Figure~\ref{flame_height_comp_FHPUF_1} to see photographic views of the bifurcated flames.  It seems that the movement of the flames as they flowed away from the fuel surface blurred out the bifurcation over time with the exception of the lower elevation areas that presented at the 90 and 100~\% contours.  Other methods of ignition for the polyurethane foam were considered in order to have a more uniform ignition of the top surface of the foam.  Each method would have resulted in a larger release of energy from the ignition source or a starter fuel.   Based on a few scoping experiments these methods did not improve the repeatability.

@@

Once the characteristics of the fires had been measured, the next step was to measure the thermal impact that the fires had on a vertical target surface.  In these experiments the target is a non-combustible wall. The non-combustible material was chosen to prevent any contribution due to combustion from the wall to the fire.  The fires were placed at predetermined, fixed positions away from the wall. The fires were prepared and conducted in the same manner as the they were in the fire characterization experiments. As a result the exposure time of the target wall to the fire was 300 seconds. In the remainder of the report the term burner will be considered synonymous with the pan for the liquid fuel or the block of solid fuel. Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and then touching the wall (0.0~m).

@@

A free standing steel frame wall with a 12.7~mm thick, 2.44~m high by 1.22~m wide, calcium silicate board face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux. Marinite Type I is the commercial name of the non-combustible board used in these experiments.  It is normally used as a fire-resistant thermal insulation that also has structural strength. 90~\% of the board is composed of calcium silicate and calcium meta-silicate.  The remaining 10~\% of the board is composed of a combination of organic fibers, fiber glass filament and crystalline silica (quartz)~\cite{BNZ:MSDS}.  Material properties that impact heat transfer were selected from the manufacturer's literature and are presented In Table~\ref{tab:Marinite_Thermal_Properties}~\cite{Marinite:1997}.
